Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-23 latex/main.tex

\section*{INTRODUCTION}
% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
% in terms of the basic goal, assumptions about samples and how recombination
% can be treated.
Classical phylogenetics is largely concerned with inferring the evolutionary history 
of species~\cite{Felsenstein2004}, with small numbers of widely diverged
samples. In this context, recombination is a nuisance which must be 
carefully accounted for~\cite{Rannala2025recombination},
affecting the inference of species trees~\cite{Posada2002effect}
and biasing downstream
analyses~\cite{Schierup2000consequences,Hedge2014bacterial}.
Modern pathogen datasets are not sparsely sampled from different species,
but consist of tens to hundreds of thousands of samples from 
a single species~\cite{
Shu2017,% GISAID ~270k FLU + others
Jolley2018open,% PubMLST, 34K genomes for e.g. Staph Auerus
Apetrei2021hiv,% HIV 1,026,217 sequences
Cryptic2022data,% TB 12k
Dyer2025enterobase% 1 million, various things like E-coli
}.
In contrast to the sophisticated model-based approaches 
used to perform species-level tree 
inference~\cite{Suchard2018bayesian,Bouckaert2019beast,Minh2020iq},
comprehensive phylogenies of these datasets are typically
built with simpler distance-matrix methods [citation?]
or approximations~\cite{Price2009fasttree}, primarily
due to computational scaling limitations, but also because of
the complexity of modelling such datasets.
While several methods have been proposed to incorporate
recombination explicitly into model-based phylogenetic 
inference~\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Muller2022bayesian}, 
scalability 
is a major issue and practical application has been limited.
Software support in explicitly \emph{using} recombination-aware
inference is also extremely limited [TODO add some concrete 
details].

The data collected during the COVID-19 pandemic as part of global
surveillance presents fundamental challenges existing appraoches
in two ways. Firstly, the sheer scale of data collected 
overwhelmed even approximate distance-matrix based methods.
At the peak in [month] 202X, there were over X thousand samples 
per day deposited in GISAID\cite{Shu2017}, with a total of 
over 17 million as of Sep 2025. % TODO check numbers
Regularly reinferring trees de-novo from rapidly 
growing datasets is untenable.
The UShER\cite{Turakhia2021} method is based on adding 
samples to an existing tree using maximum parsimony, 
with periodic rearrangements to counter the inherent 
greediness of the approach\cite{DeMaio2023}.
The benefits of this simple model and the value of 
a comprehensive genealogy updated in real-time~\cite{McBroome2021}
were conclusively demonstrated, as the resource
quickly becoming an indispensible element 
of the pandemic response\cite{Hinrichs2024}.
The second challenge posed is that recombination is 
an essential element of SARS-CoV-2 
evolution\cite{Jackson2021,VanInsberghe2021,Focosi2022,
Gutierrez2022,Ignatieva2022,Tamura2023},
with multiple deep recombination events now known to be present in 
genealogy of the global population [cites??].
Although several methods have been proposed to \emph{detect}
recombinant samples~\cite{
Turakhia2022,%RIPPLES
Zhou2023virusrecom,% CovRecomb
Alfonsi2024,%RecombinHunt 
Li2024CovRecomb%CovRecomb 
},
this information is not incorporated into the phylogenies used 
in innumerable studies~\cite{McBroome2021,Hadfield2018},
and recombination continues to be treated as a 
nuisance factor~\cite{Neches2020}.

% (e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
% we lack methods to build phylogenies that can integrate recombination
% in the face of ongoing sampling of infected cases.

% UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
% from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
% UShER has generated phylogenies updated daily \cite{McBroome2021} and
% has proven to be a powerful resource that successfully enabled the world
% to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
% These phylogenies are used to identify and monitor
% potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
% which might have acquired features to become Variants of Concern (VOC),
% for example, increased transmissibility, disease severity, and immune escape.
% Beyond supporting real-time genomic epidemiology,
% UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
% (e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
% UShER uses a tree to represent evolutionary change via mutation,
% assuming no recombination \cite{Turakhia2021}.

>>>>

\section*{INTRODUCTION}

%%% No subheadings, please. Use of abbreviations should be kept to a minimum,
%%% and nonstandard abbreviations should be defined when first used in the text.

The genomic revolution has transformed how we study the evolution and spread of pathogens,
allowing real-time tracking of pathogens over time and space.
During the COVID-19 pandemic, thousands of SARS-CoV-2 genome sequences were generated daily,
quickly overwhelming typical phylogenetic approaches.
Although real-time phylogenetic methods now exist to cope with this flood of genomic data
(e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
we lack methods to build phylogenies that can integrate recombination
in the face of ongoing sampling of infected cases.

UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
UShER has generated phylogenies updated daily \cite{McBroome2021} and
has proven to be a powerful resource that successfully enabled the world
to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
These phylogenies are used to identify and monitor
potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
which might have acquired features to become Variants of Concern (VOC),
for example, increased transmissibility, disease severity, and immune escape.
Beyond supporting real-time genomic epidemiology,
UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
(e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
UShER uses a tree to represent evolutionary change via mutation,
assuming no recombination \cite{Turakhia2021}.
However, given extensive evidence for recombination in SARS-CoV-2
\cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2023},
the tree-like representation of UShER is no longer tenable.
